Estimation of flow trajectories in a multi-lines transportation network

### 2023-02-16 latex_files/applied_network_science_article/article_v0.tex
time conflict

Case studies are divided in two sections. In the first section, we test the algorithm on toy examples, which are artifical networks where the transportation flow $n_{st}$ is randomly drawn. These examples enable some kind of validation of the algorithm, as "real" transportation flows are known and can be compared to the solutions given by our method. This setup differs from the second section, which is dedicated to applying the algorithm to the real case of the public transportation network of the city of Lausanne (tl), where embarkement and disembarkement flows are measured but transportation flows are unknown. This second case study shows that the algorithm is applicable on large, real datasets and can gives insights to transportation network planners.

@@

\subsubsection{Construction}

@@

\begin{figure}
	\label{toy_example_plots}
	\includegraphics[width=0.3\textwidth]{fig/toy_2_display.pdf}
	\includegraphics[width=0.3\textwidth]{fig/toy_3_display.pdf}
	\includegraphics[width=0.3\textwidth]{fig/toy_4_display.pdf}
	\caption{3 toy examples, with the number of tours $p \in \{2,3,4\}$. Tours are in the same color, line have a unique letter, and each stop a unique combination of a letter and a number. Clusters of nodes represent positions where transfer between tours are possible.}
\end{figure}

>>>>

Case studies are divided in two sections. In the first section, we test the algorithm on toy examples, which are artifical networks where the transportation flow $n_{st}$ is randomly drawn. These examples enable some kind of validation of the algorithm, as "real" transportation flows are known and can be compared to the solutions given by our method. This setup differs from the second section, which is dedicated to applying the algorithm to the real case of the public transportation network of the city of Lausanne (tl), where embarkement and disembarkement flows are measured but transportation flows are unknown. This second case study shows that the algorithm is applicable on large, real datasets and can gives insights about passengers probable routes in the network.

@@

\subsubsection{Examples construction}
\label{example_construction}

@@

\begin{figure}
	\includegraphics[width=0.3\textwidth]{fig/toy_2_display.pdf}
	\includegraphics[width=0.3\textwidth]{fig/toy_3_display.pdf}
	\includegraphics[width=0.3\textwidth]{fig/toy_4_display.pdf}
	\caption{3 toy examples, with the number of tours $p \in \{2,3,4\}$. Tours are displayed in the same color, line have a unique letter, and each stop a unique combination of a letter and a number. Clusters of nodes represent positions where transfer between tours are possible.}
	\label{toy_example_plots}
\end{figure}

Allowed origin-destination pairs $(s,t)$ are constructed with the same set of rules found in (REF), but without considering the exclusion criterion based on the pedestrian distance, as nodes are not spatially located. The graph shortest-path structure for the allowed origin-destination pairs is stored in the edge-trip incidence matrix $\bm{\chi}$.

A transportation flow $\mathbf{N}_\text{ref} = (n^\text{ref}_{st})$ is drawn by setting a fixed number of passengers $n^\text{ref}_{\bullet \bullet}$, and each passenger is assigned randomly to a $(s, t)$ pair drawn uniformly among all allowed origin-destination pairs. From this reference transportation flow $\mathbf{N}_\text{ref}$, using the edge-trip incidence matrix $\bm{\chi}$ and equation (\ref{equationGG}), we can compute flow on edges $\mathbf{X}_\text{ref}$ and, in turn, the number of passengers embarking $\mathbf{a}_\text{ref}$ and the number of passengers disembarking $\mathbf{b}_\text{ref}$ at each stop.

\subsubsection{Error measurements}
\label{error_measures}
Using the network structure, $\mathbf{a}_\text{ref}$, $\mathbf{b}_\text{ref}$, we can obtain a estimation of the real flow through the algorithm, noted $\mathbf{N} = (n_{st})$. There are two types of dissimilarity measure between the real data and the solution proposed by the algorithm: (1) how much $\mathbf{N}$ differs from $\mathbf{N}_\text{ref}$, and (2) how well constraints defined by the number of passengers embarking $\mathbf{a}_\text{ref}$ and disembarking $\mathbf{b}_\text{ref}$ are respected. The first dissimilarity is measured through the \emph{mean transportation error}, denoted by $\text{MTE}(\mathbf{N})$, and computed with
\begin{equation}
	\text{MTE}(\mathbf{N}) = \sum_{st} \frac{n^\text{ref}_{st}}{n^\text{ref}_{\bullet \bullet}} \frac{\lvert n_{st} - n^\text{ref}_{st}\lvert}{n^\text{ref}_{st}} = \frac{\sum_{st} \lvert n_{st} - n^\text{ref}_{st}\lvert}{n^\text{ref}_{\bullet \bullet}} 
	\label{MTE}
\end{equation} 
and the second one with the \emph{mean margin error}, noted $\text{MME}(\mathbf{N})$, obtained with
\begin{align}
	\text{MME}(\mathbf{N}) &= \frac{1}{2} \sum_{i} \frac{a^\text{ref}_i}{a^\text{ref}_\bullet} \frac{\lvert z_{\bullet i} + n_{i \bullet} - a^\text{ref}_i \lvert}{a^\text{ref}_i} + \frac{1}{2} \sum_{i} \frac{b^\text{ref}_i}{b^\text{ref}_\bullet} \frac{\lvert z_{i \bullet} + n_{\bullet i} - b^\text{ref}_i \lvert}{a^\text{ref}_i} \notag \\
	&= \frac{\sum_i (\lvert z_{\bullet i} + n_{i \bullet} - a^\text{ref}_i \lvert + \lvert z_{i \bullet} + n_{\bullet i} - b^\text{ref}_i \lvert)}{2n^\text{ref}_{\bullet \bullet}}
	\label{MME}
\end{align} 
where $z_{ij} = I((i,j) \in \mathcal{T})\sum_{st} \chi_{ij}^{st} n_{st}$ is the flow on transfer edges $\mathcal{T}$.

Note that, by construction, the mean margin error should be null when the algorithm converge. However, it can be informative to track down this error along iterations and, in some practical cases where the network is large, the algorithm convergence criterion is reached without having margin constraints perfectly respected.

\subsubsection{Algorithm iterations}
\label{example_construction}
